---
title: 'Statistical Methods for Discrete Response, Time Series, and Panel Data (W271): Lab 2'
geometry: margin=1in
output:
  github_document: default
---

# The Keeling Curve

In the 1950s, the geochemist Charles David Keeling observed a seasonal pattern in the amount of carbon dioxide present in air samples collected over the course of several years. He was able to attribute this pattern to the variation in global rates of photosynthesis throughout the year, caused by the difference in land area and vegetation cover between the Earth's northern and southern hemispheres. 

In 1958 Keeling began continuous monitoring of atmospheric carbon dioxide concentrations from the Mauna Loa Observatory in Hawaii and soon observed a trend increase carbon dioxide levels in addition to the seasonal cycle. He was able to attribute this trend increase to growth in global rates of fossil fuel combustion. This trend has continued to the present, and is known as the "Keeling Curve."

```{r load packages, echo = FALSE, message = FALSE}
library(tidyverse)
library(tsibble)
library(latex2exp)

library(patchwork)
library(lubridate)
library(tsibble)
library(feasts)
library(forecast)
library(sandwich)
library(lmtest)
library(aTSA)

library(fpp3)

theme_set(theme_minimal())
knitr::opts_chunk$set(dpi=1000)
```

```{r plot the keeling curve, echo = FALSE}
tsibble::as_tsibble(co2) %>%
  ggplot() + 
  aes(x=index, y=value) + 
  geom_line(color = 'steelblue') +
  labs(
    title = TeX(r'(Monthly Mean $CO_2$)'),
    subtitle = 'The "Keeling Curve"',
    x = 'Month and Year',
    y = TeX(r'($CO_2$ parts per million)')
  )
```
\newpage

# Your Assignment 

Your goal in this assignment is to produce a comprehensive analysis of the Mona Loa CO2 data that you will be read by an interested, supervising data scientist. Rather than this being a final report, you might think of this as being a contribution to your laboratory. You and your group have been initially charged with the task of investigating the trends of global CO2, and told that if you find "anything interesting" that the team may invest more resources into assessing the question. 

Because this is the scenario that you are responding to: 

1. Your writing needs to be clear, well-reasoned, and concise. Your peers will be reading this, and you have a reputation to maintain.
2. Decisions that you make for your analysis need also be clear and well-reasoned. While the main narrative of your deliverable might only present the modeling choices that you determine are the most appropriate, there might exist supporting materials that examine what the consequences of other choices would be. As a concrete example, if you determine that a series is an AR(1) process your main analysis might provide the results of the critical test that led you to that determination and the results of the rest of the analysis under AR(1) modeling choices. However, in an appendix or separate document that is linked in your main report, you might show what a MA model would have meant for your results instead.
3. Your code and repository are a part of the deliverable. If you were to make a clear argument that this is a question worth pursuing, but then when the team turned to continue the work they found a repository that was a jumble of coding idioms, version-ed or outdated files, and skeletons it would be a disappointment.

# Report from the Point of View of 1997 

For the first part of this task, suspend reality for a short period of time and conduct your analysis from the point of view of a data scientist doing their work in the early months of 1998. Do this by using data that is included in _every_ R implementation, the `co2` dataset. This dataset is lazily loaded with every R instance, and is stored in an object called `co2`. 

## (3 points) Task 0a: Introduction 

Introduce the question to your audience. Suppose that they _could_ be interested in the question, but they don't have a deep background in the area. What is the question that you are addressing, why is it worth addressing, and what are you going to find at the completion of your analysis. Here are a few resource that you might use to start this motivation. 

- [Wikipedia](https://en.wikipedia.org/wiki/Keeling_Curve)
- [First Publication](./background/keeling_tellus_1960.pdf)
- [Autobiography of Keeling](./background/keeling_annual_review.pdf)

## (3 points) Task 1a: CO2 data
Conduct a comprehensive Exploratory Data Analysis on the `co2` series. This should include (without being limited to) a [description of how, where and why ](https://gml.noaa.gov/ccgg/about/co2_measurements.html) the data is generated, a thorough investigation of the trend, seasonal and irregular elements. Trends both in levels and growth rates should be discussed (consider expressing longer-run growth rates as annualized averages).

What you report in the deliverable should not be your own process of discovery, but rather a guided discussion that you have constructed so that your audience can come to an understanding as succinctly and successfully as possible. This means that figures should be thoughtfully constructed and what you learn from them should be discussed in text; to the extent that there is _any_ raw output from your analysis, you should intend for people to read and interpret it, and you should write your own interpretation as well. 

```{r}
co2_tsibble <- co2 %>% as_tsibble()
```

```{r}
#EDA

## Plot 1: timeseries
plt1 <- co2_tsibble %>%
  ggplot() + 
  aes(x=index, y=value) + 
  geom_line(color = 'steelblue') +
  labs(
    title = TeX(r'(Monthly Mean $CO_2$)'),
    subtitle = 'The "Keeling Curve"',
    x = 'Month and Year',
    y = TeX(r'($CO_2$ parts per million)')
  )


# Plot 2: ACF
plt2 <- co2_tsibble %>%
  ACF(value) %>%
  autoplot() +
  labs(y = "ACF")

# Plot 3: PACF
plt3 <- ggPacf(co2_tsibble)

# Plot 4: Histogram
plt4 <- ggplot(co2_tsibble, aes(x = value)) +
  geom_histogram(bins = 15) +
  xlab(TeX(r'($CO_2$ parts per million)')) + ylab("Data Frequency")
  (plt1 + plt2)/(plt3 + plt4) + plot_annotation(
  title = (r'(Monthly Mean $CO_2$)'))

(plt1 + plt2)/(plt3 + plt4) + plot_annotation(
title = r'(Monthly Mean $CO_2$)')
```
> The top-left plot shows the timeseries over the period Jan 1959 through Jan 1997. A linear trend and annual seasonality are evident in the plot. 

> The top-right plot shows the ACF (autocorrelation function) of the series at lags of 1 month. The correlogram shows both a gradual decay and seasonality. 

> The bottom-left plot shows the PACF (partial autocorrelation function) of the series. The PACF also suggests some seasonality, due to the oscillating significant spikes at various lags. The PACF shows lags 1, 2, 12, and 13 to be significant. 

> Lastly, the bottom-right plot shows the histogram of the data and helps us understand its spread.

## (3 points) Task 2a: Linear time trend model

Fit a linear time trend model to the `co2` series, and examine the characteristics of the residuals. Compare this to a quadratic time trend model. Discuss whether a logarithmic transformation of the data would be appropriate. Fit a polynomial time trend model that incorporates seasonal dummy variables, and use this model to generate forecasts to the year 2020. 


###### Linear Time Trend
```{r}
fit_linear <- co2_tsibble %>%
  model(trend_model = TSLM(value ~ trend()))

report(fit_linear)
```
> The linear time trend models estimates an increase of 0.11 in CO2 levels per month. The model residuals are shown below. 

```{r}
# linear model residuals
fit_linear %>% gg_tsresiduals()
```

The timeseries in the top panel show the seasonality in the residuals, meaning we need to adjust the model for seasonal effects. This insight is reflected in the ACF, which shows oscillating significant lags. The histogram of the residuals is normally distributed.

```{r}
plt5 <- augment(fit_linear) %>%
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_smooth(se=FALSE, method = "loess")

plt6<- augment(fit_linear) %>%
  mutate(month = month(index, label = TRUE)) %>%
  ggplot(aes(x = month, y = .resid)) +
  geom_boxplot()

plt5 + plt6
```

> The plot in the left panel shows the residuals against the fitted values. A pattern can be clearly seen in the plot, implying heteroskedascticity in the residuals, which means that the variance in the residuals is not constant. 

> The plot in the right-panel show a box plot of residuals against the time axis. The shape of the distribution emphasizes the seasonality remaining in the data, which was not modeled by the simple linear time trend.

```{r}
augment(fit_linear) %>%
  features(.innov, ljung_box, dof = 14, lag = 24)
```
A statistically significant result from the Ljung-Box test confirms the findings from the plots: the residuals are not white noise. 

###### Quadratic Time Trend
```{r}
fit_quadratic <- co2_tsibble %>%
  model(trend_model = TSLM(value ~ trend() + I(trend()^2)))

# qudratic model residuals
fit_quadratic %>% gg_tsresiduals()
```
> The quadratic model is able to capture some of the seasonality in the data, reducing the exaggerated oscillation in the residuals in the top panel. However, the corresponding ACF still shows seasonal effects in the data. The histogram of the residuals is normal.


```{r}
plt7 <- augment(fit_quadratic) %>%
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_smooth(se=FALSE)

plt8<- augment(fit_quadratic) %>%
  mutate(month = month(index, label = TRUE)) %>%
  ggplot(aes(x = month, y = .resid)) +
  geom_boxplot()

plt7+ plt8
```

> The residuals vs fitted plot in the left panel shows evidence for heteroskedasticity. Due to the non-constant variance in the residuals, one may consider a log (or some other) transformation; however, since the original data does not show increased variance at higher levels of the data, we do not need a log transformation.

> The right panel provides further evidence for the seasonality remaining in the data. This needs to be modeled to extract back residuals resembling white noise.

```{r}
augment(fit_quadratic) %>%
  features(.innov, ljung_box, dof = 14, lag = 24)
```
> The Ljung-Box test confirms the findings from the plots: the residuals do not resemble white noise.


###### Polynomial Seasonal Time Trend
```{r}
fit_polynomial_seasonal <- co2_tsibble %>%
  model(trend_model = TSLM(value ~ trend() + I(trend()^2) + I(trend()^3) + season()))

# qudratic seasonal model residuals
fit_polynomial_seasonal %>% gg_tsresiduals()
```

> Using a polynomial seasonal time trend, we are able to remove the seasonality from the residuals: note the ACF does not show oscillating significant lags. However, we still have a case of heteroskedasticity in the residuals, shown by the non-constant variance in the top panel. The histogram shows normally distributed residuals. 

```{r}
plt9 <- augment(fit_polynomial_seasonal) %>%
  ggplot(aes(x = .fitted, y = .innov)) +
  geom_point() +
  geom_smooth(se=FALSE)

plt10<- augment(fit_polynomial_seasonal) %>%
  mutate(month = month(index, label = TRUE)) %>%
  ggplot(aes(x = month, y = .innov)) +
  geom_boxplot()

plt9 + plt10
```
> The left-panel shows that the residuals of a quadratic seasonal model still have non-constant variance, i.e. heteroskedasticity. The box plots in the right panel, however, show that including a seasonal term in the model has helped reduce the seasoality effects from the model. 

```{r}
augment(fit_polynomial_seasonal) %>%
  features(.innov, ljung_box, dof = 14, lag = 24)
```
> The Ljung-Box test shows that the residuals from the seasonal quadratic time trend model do not resemble white noise, confirming the findings from the plots shown above. 

```{r}
# generate forecasts using the seasonal model

fit_polynomial_seasonal %>%
  fabletools::forecast(h=36) %>%
  autoplot(co2_tsibble)
```

> The polynomial seasonal model produces forecasts that extrapolate the linear trend of the series and include the seasonality of the data. However, as noted in the sections above, the residuals of this model do not represent white noise. Hencel, a better model may be possible for this data. 

## (3 points) Task 3a: ARIMA times series model 

Following all appropriate steps, choose an ARIMA model to fit to the series. Discuss the characteristics of your model and how you selected between alternative ARIMA specifications. Use your model (or models) to generate forecasts to the year 2022. 

> We can begin by evaluating the need for differencing in the data. The timeseries is non-stationary and hence first-order differencing may be required. The unitroot_kpss tests below provide this evaluation objectively:

```{r}
unitroot_kpss(co2_tsibble$value)
```

> Without any differencing, the `unitroot_kpss` function shows that the series is not stationary.


```{r}
unitroot_kpss(co2_tsibble$value %>% diff())
```


> After first-order differencing, however, we can see that the test fails to reject the null hypothesis (the series is stationary) at the 0.05% significance level. This result is confirmed by the function unitroot_ndiffs():

```{r}
unitroot_ndiffs(co2_tsibble$value)
```

> The residuals of the differenced series are shown below. 

```{r}
co2_tsibble %>% gg_tsdisplay(difference(value), plot_type="partial") +labs(subtitle = "CO2 levels")
```
> The ACF exhibits seasonality in the residuals, implying that we may need seasonal differencing. 

```{r}
seasonally_differenced_arima <- co2_tsibble %>% model(seasonal_ARIMA = 
                                                        ARIMA(value~0+pdq(0,1,0) + PDQ(0,1,0)))
                                                        # include non-seasonal differencing as well

seasonally_differenced_arima %>% residuals() %>% gg_tsdisplay(y = .resid, plot_type="partial")
```

> Including seasonal as well as non-seasonal differencing in the model produces residuals much close to resembling white noise, as is seen in the top panel. The ACF plot shows significance at lags 1 and 3 and higher lags of 9, 11, and 12. The PACF also shows significance at lags 1 and 3 and some of the higher lags. 

> We can attempt a ARIMA(1,1,1)(0,1,0)[12] model. Note the [12] in model notation indicates the seasonality in the data.

```{r}
seasonally_differenced2_arima <- co2_tsibble %>% model(seasonal_ARIMA = 
                                                        ARIMA(value~0+pdq(1,1,1) + PDQ(0,1,0)))
                                                        # include non-seasonal differencing as well

seasonally_differenced2_arima %>% residuals() %>% gg_tsdisplay(y = .resid, plot_type="partial")

```

> Including an AR(1) term and an MA(1) term reduces the number of significant lags in both the ACF and the PACF. However, we still have a significant lag at lag 12. We can use attempt another model with an additional term: P=1.


```{r}
seasonally_differenced3_arima <- co2_tsibble %>% model(seasonal_ARIMA = 
                                                        ARIMA(value~0+pdq(1,1,1) + PDQ(1,1,0)))
                                                        # include non-seasonal differencing as well

seasonally_differenced3_arima %>% residuals() %>% gg_tsdisplay(y = .resid, plot_type="partial")
```

> The additional term does not completely remove the significant lag 12 from the ACF and PACF. Additionally, the lag 24 remains significant in both plots.

> We can now use auto-ARIMA to help us find the optimal model.

```{r}
model_fit <- co2_tsibble %>% model(ARIMA_fit = ARIMA(value))

report(model_fit[1])
```
> Residual Diagnostics

```{r}
model_fit %>% residuals() %>% gg_tsdisplay(y = .resid, plot_type="partial")
```

> The residual plots indicate resemblance to white noise. We can check this conclusion with the Ljung-Box test.

```{r}
augment(model_fit) %>%
  features(.innov, ljung_box, dof = 14, lag = 24)
```

> Unfortunately, the Ljung-Box test still rejects the null hypothesis of stationarity. We can also use the ADF test to check for residual stationarity.

```{r}
adf.test(model_fit %>% residuals %>% select(.resid) %>% as_vector(), alternative = "stationary", k=5)
```
> The ADF test also suggests non-stationarity in the residuals. While the residuals are close to resembling white noise, they are not completely white noise. There still may be additional information in the residuals that can be modeled out. However, the optimal model identified by auto-ARIMA is already far from being parsimonius: ARIMA(1,1,1)(1,1,2)[12]. Adding more parameters will increase the complexity of the model with a relatively little gain in moving the residuals further towards resembling white noise.

## (3 points) Task 4a: Forecast atmospheric CO2 growth 

Generate predictions for when atmospheric CO2 is expected to be at [420 ppm](https://research.noaa.gov/article/ArtMID/587/ArticleID/2764/Coronavirus-response-barely-slows-rising-carbon-dioxide) and 500 ppm levels for the first and final times (consider prediction intervals as well as point estimates in your answer). Generate a prediction for atmospheric CO2 levels in the year 2100. How confident are you that these will be accurate predictions?

```{r}
forecasts <- model_fit %>% fabletools::forecast(h=1248)

forecasts %>%
  autoplot(co2_tsibble)
```

```{r}
data.frame(forecasts) %>% filter(.mean >= 419.5 & .mean <= 420.5) %>% arrange(index)
```

> The global CO2 levels, as per the model estimated above, are first expected to reach 420 ppm in April 2032. And the final time for the CO2 levels to reach 420 ppm will be in Oct, 2035.

```{r}
data.frame(forecasts) %>% filter(.mean >= 499.5 & .mean <= 500.5) %>% arrange(index)
```

> The global CO2 levels, as per the model estimated above, are first expected to reach 500 ppm in April 2084. And the final time for the CO2 levels to reach 500 ppm will be in Aug, 2086.

> TODO: How confident are you in the predictions? Extract the prediction intervals for this answer.

# Report from the Point of View of the Present 

One of the very interesting features of Keeling and colleagues' research is that they were able to evaluate, and re-evaluate the data as new series of measurements were released. This permitted the evaluation of previous models' performance and a much more difficult question: If their models' predictions were "off" was this the result of a failure of the model, or a change in the system? 

## (1 point) Task 0b: Introduction 

In this introduction, you can assume that your reader will have **just** read your 1997 report. In this introduction, **very** briefly pose the question that you are evaluating, and describe what (if anything) has changed in the data generating process between 1997 and the present. 

## (3 points) Task 1b: Create a modern data pipeline for Mona Loa CO2 data.

The most current data is provided by the United States' National Oceanic and Atmospheric Administration, on a data page [[here](https://gml.noaa.gov/ccgg/trends/data.html)]. Gather the most recent weekly data from this page. (A group that is interested in even more data management might choose to work with the [hourly data](https://gml.noaa.gov/aftp/data/trace_gases/co2/in-situ/surface/mlo/co2_mlo_surface-insitu_1_ccgg_HourlyData.txt).) 

Create a data pipeline that starts by reading from the appropriate URL, and ends by saving an object called `co2_present` that is a suitable time series object. 

Conduct the same EDA on this data. Describe how the Keeling Curve evolved from 1997 to the present, noting where the series seems to be following similar trends to the series that you "evaluated in 1997" and where the series seems to be following different trends. This EDA can use the same, or very similar tools and views as you provided in your 1997 report. 

## (1 point) Task 2b: Compare linear model forecasts against realized CO2

Descriptively compare realized atmospheric CO2 levels to those predicted by your forecast from a linear time model in 1997 (i.e. "Task 2a"). (You do not need to run any formal tests for this task.) 

## (1 point) Task 3b: Compare ARIMA models forecasts against realized CO2  

Descriptively compare realized atmospheric CO2 levels to those predicted by your forecast from the ARIMA model that you fitted in 1997 (i.e. "Task 3a"). Describe how the Keeling Curve evolved from 1997 to the present. 

## (3 points) Task 4b: Evaluate the performance of 1997 linear and ARIMA models 

In 1997 you made predictions about the first time that CO2 would cross 420 ppm. How close were your models to the truth? 

After reflecting on your performance on this threshold-prediction task, continue to use the weekly data to generate a month-average series from 1997 to the present, and compare the overall forecasting performance of your models from Parts 2a and 3b over the entire period. (You should conduct formal tests for this task.) 

## (4 points) Task 5b: Train best models on present data

Seasonally adjust the weekly NOAA data, and split both seasonally-adjusted (SA) and non-seasonally-adjusted (NSA) series into training and test sets, using the last two years of observations as the test sets. For both SA and NSA series, fit ARIMA models using all appropriate steps. Measure and discuss how your models perform in-sample and (psuedo-) out-of-sample, comparing candidate models and explaining your choice. In addition, fit a polynomial time-trend model to the seasonally-adjusted series and compare its performance to that of your ARIMA model.

## (3 points) Task Part 6b: How bad could it get?

With the non-seasonally adjusted data series, generate predictions for when atmospheric CO2 is expected to be at 420 ppm and 500 ppm levels for the first and final times (consider prediction intervals as well as point estimates in your answer). Generate a prediction for atmospheric CO2 levels in the year 2122. How confident are you that these will be accurate predictions?
